# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

This assignment utilizes the source data file from:

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

Although the file is already available within the repository, first check whether `activity.csv` file exists. If not, the code:

- Downloads a new zip file into a local copy with **today's date**
- Unzips the file and reads it into a dataframe.

Given the appropriate csv, use `read.csv()` to read file. Finally, convert date string into a date format for future manipulation.

```{r}
options(scipen=999) # Disable scientific notation in output
if (!file.exists("activity.csv")) {
        zipfileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        zipfilename <- paste0("rep-data-activity-",strftime(Sys.time(),"%Y%m%d"),".zip")
        download.file(url=zipfileurl,destfile=zipfilename,method="curl")
        unzip(zipfile=zipfilename)
}

activity <- read.csv("activity.csv",stringsAsFactors=FALSE)
activity$date <- as.Date(activity$date)

```

## What is mean total number of steps taken per day?

Create a new dataframe for the steps per day with the `ddply` from the `plyr` package. The results are a summarized dataframe with the total steps per day.

```{r}
library(plyr)
stepsperday <- ddply(activity, .(date),summarise,steps=sum(steps))

```

Plot the histogram of the number of steps per day. A bin width of a $max\:steps\over{20}$ was selected after various exploratory graphs.

```{r}
library(ggplot2)
ggplot(stepsperday,aes(steps)) + 
        geom_histogram(binwidth=max(stepsperday$steps,na.rm=TRUE)/20) +
        labs(title="Histogram of Steps Per Day",x="Steps",y="Frequency")
````

Find the mean and median. Since the data includes a number of missing observations, utilize `na.rm=TRUE`

```{r}
meanperday <- mean(stepsperday$steps,na.rm=TRUE)
medianperday <- median(stepsperday$steps,na.rm=TRUE)
meanperday
medianperday
```

The mean is **`r meanperday`** and the median is **`r medianperday`**.

## What is the average daily activity pattern?

Leverage `ddply` again, this time to average the steps per interval. Next `which.max` simplifies finding the interval with the highest average. For the plot, the blue lines highlight the location of the max and its interval.

```{r}
stepsperinterval <- ddply(activity,.(interval),summarise,steps=mean(steps,na.rm=TRUE))
indexofmax <- which.max(stepsperinterval$steps)
maxinterval <- stepsperinterval[indexofmax,"interval"]
maxsteps <- stepsperinterval[indexofmax,"steps"]

ggplot(stepsperinterval,aes(interval,steps)) +
        geom_line() +
        geom_hline(aes(yintercept=maxsteps),color="blue") +
        geom_vline(xintercept=maxinterval,color="blue") +
        labs(title="Average Steps Per Interval",x="Interval",y="Steps")
```

The interval with the greatest average number of steps was **`r maxinterval`** with **`r maxsteps`** steps.

## Imputing missing values

First, find the number of observations with missing values (**`NA`**)

```{r}
missingcount <- sum(!complete.cases(activity))
```

With `r missingcount` missing observations, let's see if there is a pattern. Summarising the missing values by date provides:


```{r}
missingintervalcount <- ddply(activity,.(date),summarise,missing=sum(is.na(steps)))
missingintervalcount[missingintervalcount$missing>0,]

```

Given that there are 288 five minute intervals in a day - the pattern of missing data is all the data on a given day. Perhaps imputing the missing intervals with the average intervals for the same day of the week for the same interval is a good approach since individual movement patterns tend to be cyclical by week. First, assign the week day to date column and summarize the missing steps by the week day. 

```{r}
activity$day <- weekdays.Date(activity$date)
ddply(activity,.(day),summarise,missing=sum(is.na(steps)))
```

This shows us that at most, 2 days are missing for any given week day with Tuesday not having any missing data. Next, create a new data frame with means for each interval and each week day. This is used to merge into the new `imputed` data frame.

```{r}
weekdaymeans <- ddply(activity,.(day,interval),summarise,meanstepweekday=mean(steps,na.rm=TRUE))
imputed <- merge(activity,weekdaymeans,by=c("day","interval"))
imputed$steps[is.na(imputed$steps)] <- imputed$meanstepweekday[is.na(imputed$steps)]
imputed$meanstepweekday <- NULL

```

Now, a new histogram, mean and median with the `imputed` activity data.

```{r}

imputedstepsperday <- ddply(imputed, .(date),summarise,steps=sum(steps))

library(ggplot2)
ggplot(imputedstepsperday,aes(steps)) + 
        geom_histogram(binwidth=max(imputedstepsperday$steps,na.rm=TRUE)/20) +
        labs(title="Histogram of Steps Per Day (imputed data)",x="Steps",y="Frequency")
````

```{r}
imputedmeanperday <- mean(imputedstepsperday$steps)
imputedmedianperday <- median(imputedstepsperday$steps)
imputedmeanperday
imputedmedianperday
```

The mean of the imputed data is **`r imputedmeanperday`** and the median is **`r imputedmedianperday`**. Compared to the results from the original dataset, removing the `NA` had a mean of **`r meanperday`** and a median of **`r medianperday`** - so both the mean and median slightly after imputing the data with this method.

## Are there differences in activity patterns between weekdays and weekends?

Create a new `daytype` variable to categorize weekday versus weekend and use this to facet a plot to find patterns among intervals. Lines are added to represent the mean and max of the respective sets of weekend versus weekday data to highlight differences.

```{r}
imputed$daytype <- as.factor(ifelse(weekdays(imputed$date) %in% c("Saturday","Sunday"), "weekend", "weekday"))
stepsperimpinterval <- ddply(imputed,.(interval,daytype),summarise,steps=mean(steps,na.rm=TRUE))

ggplot(stepsperimpinterval,aes(interval,steps)) +
        geom_line() +
        facet_wrap(~ daytype,nrow=2) +
        geom_line(stat="hline", linetype="dotted", yintercept="mean") +
        geom_line(stat="hline", linetype="dotted", yintercept="max") +
        labs(title="Average Steps Per Interval",x="Interval",y="Steps")
```


